---
title: "tana_mortality_IFR"
author: "Benny Rice"
date: "8/02/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE, message = FALSE, warning = FALSE)
```

***
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

## HEADER INFO

**Metcalf Lab | Department of Ecology and Evolutionary Biology | Princeton University**

- July 2020
- Code associated with project aimed at detecting mortality anomalies in Tana
- Apologies in advance for amateurish code (ie all the loops and clunkiness)

</div>

***

##### (i) SET UP

Loading necessary packages
```{r i1, echo = FALSE, eval = TRUE, message = FALSE}
library(tidyverse)
library(kableExtra)   #For printing tables in R markdown
library(knitr)        #For printing tables in R markdown
library(reshape2)     #Melting data frames as needed
library(ggrepel)      #Plotting country names neatly
library(cowplot)      #ggplot add on for better plots
library(patchwork)    #Combining ggplots neatly
library(mgcv)         #GAMs
library(here)         #Finding files easily
```

Reading in data sets:

1. IFR estimates: `ifrs_from_lit.csv`
2. UN age data: `un_ages.csv`
3. Tana age data: `TNR_age.csv`

```{r i2, echo = FALSE, eval = TRUE}
#1
ifr_ests <- read_csv(here("ifrs_from_lit.csv"))
#2
un_ages <- read_csv(here("un_ages.csv"))
#3
df.tana <- read_csv(here("TNR_age.csv"))
```


***
***

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

### Expected mortality burden under different IFR scenarios

</div>

***
***

##### Section 1: Making age vs IFR curves: Fitting age to IFR estimates

***

- Fitting age to IFR estimates from France, Italy, China using a GAM

```{r 1.01, echo = FALSE, eval = TRUE}
# Fit age to IFR estimates ------------------------------------------------
ifr_ests$predictor <- (ifr_ests$age_lower + ifr_ests$age_upper)/2
ifr_ests$ifr_prop_est <- ifr_ests$ifr_perc_est/100

ifr_fit_age <- gam(ifr_prop_est ~ s(predictor), family = betar(link="logit"), 
               data = ifr_ests, method = "REML")

ages <- seq(0, 100, by = 1)

ifr_preds_age <- predict(ifr_fit_age,  data.frame(predictor = ages), 
                     type = "response", se.fit = TRUE)
ifrs_to_plot_age <- data.frame(ages = ages,
                               est = ifr_preds_age$fit, 
                               upper = ifr_preds_age$fit + 2*ifr_preds_age$se.fit, 
                               lower = ifr_preds_age$fit - 2*ifr_preds_age$se.fit)
```

***

##### Section 2: Plotting IFR curves 

***

- Writing a function to fit and shift IFR curves

```{r 2.01, echo = FALSE, eval = TRUE}
predict_ifr <- function(predictor, gam = ifr_fit_age) {
  predict(gam, newdata = data.frame(predictor = predictor), type = "response")
}
```

- Plotting baseline and shifted IFR vs age curves

```{r 2.02, echo = FALSE, eval = TRUE}
#Creating a ifr by age table for plotting
age_ifrs <- data.frame(age = seq(1, 100, by = 1))
age_ifrs %>%
  mutate(ifr_by_age_plus00y = predict_ifr(predictor = age + 0,      gam = ifr_fit_age), 
         ifr_by_age_plus02y = predict_ifr(predictor = age + 2,  gam = ifr_fit_age),
         ifr_by_age_plus04y = predict_ifr(predictor = age + 4,  gam = ifr_fit_age),
         ifr_by_age_plus06y = predict_ifr(predictor = age + 6,  gam = ifr_fit_age),
         ifr_by_age_plus08y = predict_ifr(predictor = age + 8,  gam = ifr_fit_age),
         ifr_by_age_plus10y = predict_ifr(predictor = age + 10, gam = ifr_fit_age)) %>%
  pivot_longer(starts_with("ifr"), names_to = "ifr_type", values_to = "ifr_est") %>%
  mutate(ifr_type = factor(ifr_type, levels = c("ifr_by_age_plus00y", 
                                                "ifr_by_age_plus02y", 
                                                "ifr_by_age_plus04y", 
                                                "ifr_by_age_plus06y", 
                                                "ifr_by_age_plus08y", 
                                                "ifr_by_age_plus10y"))) -> ages_to_plot

#Setting colors
ifr_cols <- c("ifr_by_age_plus00y" = "#ffccbc", 
              "ifr_by_age_plus02y" = "#ff8a65", 
              "ifr_by_age_plus04y" = "#ff7043", 
              "ifr_by_age_plus06y" = "#ff5722", 
              "ifr_by_age_plus08y" = "#e64a19", 
              "ifr_by_age_plus10y" = "#bf360c")
ifr_labs <- c("Baseline", 
              "Shifted (2y younger)", 
              "Shifted (4y younger)", 
              "Shifted (6y younger)", 
              "Shifted (8y younger)", 
              "Shifted (10y younger)")
names(ifr_labs) <- names(ifr_cols)

#Plotting
ggplot(data = ages_to_plot, aes(x = age, y = ifr_est, group = ifr_type)) +
  geom_line(aes(color = ifr_type)) + 
  scale_color_manual(values = ifr_cols, labels = ifr_labs,
                     name = "Predictor of IFR") +
  theme_minimal_hgrid(color = "white") +
  labs(x = "Age", y = "IFR") +
  theme(legend.position = "right",
        text = element_text(size = 12))
```

***

##### Section 3: Sourcing demography age structure data

***

3.1 | **UN age data:** Madagascar country wide age pattern as a placeholder until get better Tana-specific data

- Source: UNPOP 2020 age data
- Calculating proportion of population in each age group

```{r 3.01, echo = FALSE, eval = TRUE}
mada.ages <- un_ages %>% filter(country == "Madagascar") %>% select(age, pop) %>% mutate(prop = pop/(sum(pop))) %>%
  #Binning all ages over 95 such that 95 is the last age (corresponds to ages 95+)
  mutate(prop = if_else(age == 95, sum(prop[ages >= 95]), prop)) %>% filter(age <= 95)
```

***

3.2 | **Tana data:** Using data from draft of 2018 census (?)

- Source: Fidy via Tanjona via CCO?

```{r 3.02, echo = FALSE, eval = TRUE}
tana.ages <- df.tana %>% mutate(age = Age) %>% select(age, pop) %>% mutate(prop = pop/(sum(pop)))
```

***

3.3 | **WorldPop data:** Using X boundary around Tana

(in progress)


***

- Comparing demography data sources: UN Madagascar age structure data vs Tana 2018 census data

```{r 3.04, echo = FALSE, eval = TRUE}
full_join(mada.ages, tana.ages, by = "age") %>%
  rename(mada.prop = prop.x) %>% rename(tana.prop = prop.y) %>% select(age, mada.prop, tana.prop) %>%
  melt(id = c("age")) %>%
  ggplot(aes(x = age, y = value, color = variable)) +
    geom_point() +
    geom_smooth() +
    ylab("Proportion of the population") + xlab("Age") +
    theme_bw()
```

***

##### Section 4: Infection over age pattern scenarios

***

4.1 | **Baseline:** Infections are randomly distributed among age groups

  - (ie the infected population reflects the age structure of the population as a whole)

```{r 4.01, echo = FALSE, eval = TRUE}
#Creating a data frame showing constant probability of infection among age groups
inf.even <- data.frame(age = 0:95, pr_inf = rep(1, 96))
```

***

***

4.2 | **Susceptibility:** Susceptibility to infection varies by age groups

  - (ie the infected population reflects age specific differences in susceptibility)
  - (ie exposure is equally likely but the probability of infection upon exposure is reduced for children)
  - Data sourced from: [Davies et al 2020 Nature Medicine](https://www.nature.com/articles/s41591-020-0962-9)
    - Davies, N.G., Klepac, P., Liu, Y. et al. Age-dependent effects in the transmission and control of COVID-19 epidemics. Nat Med (2020). https://doi.org/10.1038/s41591-020-0962-9
    - Data taken from: Extended Data Fig. 4: Posterior estimates for the consensus susceptibility and clinical fraction from 6 countries 

Davies et al mean age specific susceptibility:

| Age Group | Susc |
| --------- | ---- |
| 0-9       | 0.40 |


```{r 4.02, echo = FALSE, eval = TRUE}
#Creating a data frame showing constant probability of infection among age groups
inf.even <- data.frame(age = 0:95, pr_inf = rep(1, 96))
```

***


***

##### Section 5: Writing a function to return an expected number of deaths

***

Function inputs
1. A given number of infected individuals
2. A specified age structure (ie the porportion of the population in each age group)
3. Number of years to shift the IFR curve younger (from the GAM-fit-to-Verity-etc baseline)

***

- Note: The probability of death for one infection (`n_infs = 1`) can be multiplied by the number of infections to give the total number of deaths expected: `D = P(death|inf) * n_infs`

```{r 5.01, echo = FALSE, eval = FALSE}
f.estimate_deaths <- function(n_infs = 1, demo.age.str = tana.ages, ifr_shift = 0, inf.age.str = inf.even){
  
  
  n_infs <- 1
  age_str_data <- tana.ages
  ifr_shift <- 0
  inf.age.str <- inf.even
  
  #pop_i the number of infected individuals expected for each age group
  df.pop <- age_str_data %>% mutate(pop_i = prop*n_infs)
  #Calculating the number of deaths expected per age for the specified IFR scenario
  expected_deaths <- df.pop %>%
    mutate(ifr_by_ageplusx = predict_ifr(predictor = age + ifr_shift, gam = ifr_fit_age)) %>%
    pivot_longer(starts_with("ifr"), names_to = "ifr_type", values_to = "ifr_est") %>%
    mutate(deaths = ifr_est*pop_i)
  #Summing the number of total number of deaths
  total_deaths <- sum(expected_deaths$deaths)
  
  return(total_deaths)
}
```

```{r 5.02, echo = FALSE, eval = FALSE}



v.ages <- c("Age_0_9", "Age_10_19", "Age_20_29", "Age_30_39", "Age_40_49", "Age_50_59", "Age_60_69", "Age_70plus")
v.susc <- c(0.40, 0.38, 0.79, 0.86, 0.80, 0.82, 0.88, 0.74)

dfx <- data.frame(age_group = v.ages, susc = v.susc, stringsAsFactors = FALSE) %>% 
  mutate(rel.susc = susc/sum(susc))

sum(dfx$rel.susc)


dfx %>% ggplot(aes(x = age_group, y = rel.susc)) + 
  geom_point() +
  #geom_hline(yintercept = 1) +
  xlab("Age group") + ylab("Rel. Susceptibility") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

dfx %>% ggplot(aes(x = age_group, y = rel.susc)) + 
  geom_bar(position="dodge", stat="identity") +
  #geom_hline(yintercept = 1) +
  xlab("Age group") + ylab("Susceptibility") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))
```



```{r 5.01X, echo = FALSE, eval = FALSE}
f.estimate_deaths(n_infs = 0, age_str_data = tana.ages, ifr_shift = 0)
f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 0)
f.estimate_deaths(n_infs = 2, age_str_data = tana.ages, ifr_shift = 0)

f.estimate_deaths(n_infs = 0, age_str_data = tana.ages, ifr_shift = 10)
f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 10)
f.estimate_deaths(n_infs = 2, age_str_data = tana.ages, ifr_shift = 10)




#For the first plot:
# Calculating the expected number of deaths for a given number of infections
#   Because the increase is linear, can simply calculate the expected number of deaths for 1 infection 
#     then multiply by the number of infections: Deaths = P(death|inf) * n_infs

data.frame(n_inf = 1:50000) %>% 
  mutate(m.ifr_plus00y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 0),
         m.ifr_plus02y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 2),
         m.ifr_plus04y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 4),
         m.ifr_plus06y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 6),
         m.ifr_plus08y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 8),
         m.ifr_plus10y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 10)) %>%
  melt(id = "n_inf") %>%
  ggplot(aes(x = n_inf, y = value, color = variable)) +
    geom_point() +
    scale_color_manual(values = c("#ffccbc", "#ff8a65", "#ff7043", "#ff5722", "#e64a19", "#bf360c"),
                       name = "IFR scenario",
                       labels = c("Baseline", "Shifted (2y younger)", "Shifted (4y younger)", 
                                  "Shifted (6y younger)", "Shifted (8y younger)", "Shifted (10y younger)")) +
    geom_hline(yintercept = 65.5, color = "#00bfa5") +
    ylab("Expected weekly mortality") +
    theme_bw()

data.frame(n_inf = 1:50000) %>% 
  mutate(m.ifr_plus00y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 0),
         m.ifr_plus02y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 2),
         m.ifr_plus04y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 4),
         m.ifr_plus06y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 6),
         m.ifr_plus08y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 8),
         m.ifr_plus10y = n_inf*f.estimate_deaths(n_infs = 1, age_str_data = tana.ages, ifr_shift = 10)) %>%
  melt(id = "n_inf") %>%
  filter(value > 65.5) %>% group_by(variable) %>%
  summarize(D.n_inf = min(n_inf), .groups = "drop")
```






